---
title: "Final Project: Statistical Learning"
author: "María Colado Montañés and Luis Calderón Yunda"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
```

```{r packages, echo=FALSE}
# Core packages
library(tidyverse)
library(janitor)   # clean_names() — same as Midterm 1
library(e1071)
library(caret)
library(pROC)
library(kernlab)
library(class)     # for KNN
library(tree)      # for tree
library(rpart)
library(randomForest)
library(gbm)
library(xgboost)

# Visualization
library(ggplot2)
library(patchwork)
library(viridis)
library(scales)

set.seed(42)
```

# Introduction

In this part of the project, we focus on KNN, Tree-Based Methods and Support Vector Machines (SVMs). Unlike probabilistic classifiers, the latter are margin-based methods that aim to find the decision boundary that best separates the classes, while allowing the use of kernel functions to capture possible non-linear structure.

Our analysis builds directly on the results of Midterm 1. In particular, we use the previous evidence on class overlap, non-linearity and class imbalance to justify specifically the SVM models fitted here and the evaluation strategy adopted.

```{r data-loading, echo=FALSE}
# Using relative paths for reproducibility
df_raw <- read_csv("data/Customer-Churn-Records.csv") %>%
  clean_names()

df <- df_raw %>%
  dplyr::select(-any_of(c("row_number", "customer_id", "surname", "complain"))) %>%
  mutate(
    exited           = factor(exited, levels = c(0, 1), labels = c("No", "Yes")),
    geography        = as.factor(geography),
    gender           = as.factor(gender),
    has_cr_card      = factor(has_cr_card,      levels = c(0, 1), labels = c("No", "Yes")),
    is_active_member = factor(is_active_member, levels = c(0, 1), labels = c("No", "Yes")),
    card_type        = factor(card_type)
  )

cat("Dimensions:", nrow(df), "x", ncol(df), "\n")
cat("Class distribution:\n")
print(prop.table(table(df$exited)))
```

## Train/test split

We use the same 80/20 stratified split as in Midterm 1 to ensure comparability of results.

```{r train-test-split}
set.seed(42)
train_idx <- createDataPartition(df$exited, p = 0.8, list = FALSE)
train_data <- df[train_idx, ]
test_data  <- df[-train_idx, ]
```

## Cost structure and evaluation framework

We retain the asymmetric cost structure defined broadly in this project:

- **False Negative** (missed churner): cost = **100**
- **False Positive** (loyal customer incorrectly flagged): cost = **10**

The theoretical cost-sensitive threshold is therefore: $\pi^* = \frac{C_{FP}}{C_{FP} + C_{FN}} = \frac{10}{10 + 100} \approx 0.09$.

As in others parts of the project, we use **AUC** as the main threshold-independent criterion to compare models and tune hyperparameters, since it evaluates how well each classifier ranks churners above non-churners regardless of the classification cutoff. However, because the bank’s decision problem is asymmetric, the final practical assessment is based on **Expected Cost**, supplemented by Sensitivity, Specificity, Precision and F1.

```{r cost-helpers}
# ── Cost and evaluation helpers ──────────────────────────────────────────────
FN_cost  <- 100
FP_cost  <- 10
pi_star  <- FP_cost / (FP_cost + FN_cost)   # ≈ 0.09

expected_cost <- function(actual, predicted) {
  TP <- sum(actual == "Yes" & predicted == "Yes")
  TN <- sum(actual == "No"  & predicted == "No")
  FP <- sum(actual == "No"  & predicted == "Yes")
  FN <- sum(actual == "Yes" & predicted == "No")
  (FP * FP_cost + FN * FN_cost) / length(actual)
}

evaluate_model <- function(actual, probs, model_name, threshold = pi_star) {
  predicted <- factor(if_else(probs >= threshold, "Yes", "No"),
                      levels = c("No", "Yes"))
  cm  <- confusionMatrix(predicted, actual, positive = "Yes")
  roc_obj <- roc(actual, probs, levels = c("No", "Yes"), quiet = TRUE)

  tibble(
    Model       = model_name,
    Threshold   = threshold,
    AUC         = round(as.numeric(auc(roc_obj)), 3),
    Accuracy    = round(cm$overall["Accuracy"],       3),
    Sensitivity = round(cm$byClass["Sensitivity"],    3),
    Specificity = round(cm$byClass["Specificity"],    3),
    Precision   = round(cm$byClass["Pos Pred Value"], 3),
    F1          = round(cm$byClass["F1"],             3),
    Exp_Cost    = round(expected_cost(actual, predicted), 3)
  )
}

# Threshold sweep: evaluates sensitivity, specificity and expected cost across thresholds
sweep_threshold <- function(actual, probs, thresholds = seq(0.01, 0.99, by = 0.01)) {
  map_dfr(thresholds, function(th) {
    pred <- factor(if_else(probs >= th, "Yes", "No"), levels = c("No", "Yes"))
    tibble(
      threshold   = th,
      sensitivity = sum(actual == "Yes" & pred == "Yes") / sum(actual == "Yes"),
      specificity = sum(actual == "No"  & pred == "No")  / sum(actual == "No"),
      exp_cost    = expected_cost(actual, pred)
    )
  })
}
```

## Modelling strategy

### Excluding the Maximal Margin Classifier

The maximal margin classifier requires the classes to be perfectly linearly separable. This assumption is not plausible in our churn dataset. In Midterm 1, both the visual analysis and the results of the probabilistic classifiers showed substantial overlap between churners and non-churners, with no clean linear separation. In particular, LDA achieved only moderate discrimination, while QDA suggested that the decision boundary is better understood as non-linear. Attempting to fit a hard-margin SVM would cause the solver to fail or return a degenerate result.

For this reason, we do not consider the Maximal Margin Classifier in this report. Instead, we move directly to the **Support Vector Classifier** (soft margin), which allows classification errors, and to the **kernel SVM**, which also relaxes the linearity assumption.

### Training on the original imbalanced data

In Midterm 1, we compared no resampling, random oversampling and SMOTE across the probabilistic classifiers. The main conclusion was that resampling did not provide a clear advantage for the more flexible models. In particular, QDA achieved both its best AUC and its lowest expected cost when trained on the original imbalanced data.

This result is especially relevant here because the RBF SVM is also a flexible non-linear classifier. For this reason, we train all SVM models on the original data and handle the asymmetry of the problem through the cost-sensitive threshold, as in our previous work. This keeps the analysis consistent across both parts of the project and avoids introducing synthetic observations unless they are clearly beneficial.

```{r evaluate-ensemble, echo = FALSE}
evaluate_model_xgb <- function(preds, true_labels, probs) {

  cm <- confusionMatrix(preds, true_labels, positive = "Yes")
  
  # Expected Cost Calculation (FN = 100, FP = 10)
  fn <- sum(preds == "No"  & true_labels == "Yes")
  fp <- sum(preds == "Yes" & true_labels == "No")
  exp_cost <- (fn * 100 + fp * 10) / length(true_labels)
  
  roc_obj <- roc(response = true_labels, predictor = probs, quiet = TRUE)
  
  tibble(
    AUC           = as.numeric(auc(roc_obj)),
    Accuracy      = cm$overall["Accuracy"],
    Sensitivity   = cm$byClass["Sensitivity"],
    Specificity   = cm$byClass["Specificity"],
    Precision     = cm$byClass["Precision"],
    F1_Score      = cm$byClass["F1"],
    Expected_Cost = exp_cost
  )
}

evaluate_knn <- function(pred_knn, true_labels, threshold = 0.5,
                         cost_fn = 100, cost_fp = 10,
                         positive_class = "Yes") {
  
  raw_prob <- attr(pred_knn, "prob")
  
  if (is.null(raw_prob)) {
    stop("The pred_knn object does not have probabilities. You must fit knn(..., prob = TRUE).")
  }
  
  true_labels <- factor(true_labels, levels = c("No", "Yes"))
  pred_knn    <- factor(pred_knn, levels = c("No", "Yes"))
  
  prob_yes <- ifelse(pred_knn == positive_class, raw_prob, 1 - raw_prob)
  
  preds_thr <- factor(
    ifelse(prob_yes >= threshold, "Yes", "No"),
    levels = c("No", "Yes")
  )
  
  cm <- caret::confusionMatrix(preds_thr, true_labels, positive = positive_class)
  
  fn <- sum(preds_thr == "No"  & true_labels == "Yes")
  fp <- sum(preds_thr == "Yes" & true_labels == "No")
  exp_cost <- (fn * cost_fn + fp * cost_fp) / length(true_labels)
  
  roc_obj <- pROC::roc(response = true_labels, predictor = prob_yes, quiet = TRUE)
  
  tibble::tibble(
    AUC           = as.numeric(pROC::auc(roc_obj)),
    Accuracy      = unname(cm$overall["Accuracy"]),
    Sensitivity   = unname(cm$byClass["Sensitivity"]),
    Specificity   = unname(cm$byClass["Specificity"]),
    Precision     = unname(cm$byClass["Precision"]),
    F1_Score      = unname(cm$byClass["F1"]),
    Expected_Cost = exp_cost
  )
}

thresh_theo <- 0.09
my_colors <- c("No" = "#00BFFF", "Yes" = "#FF7F50")
```

# Support Vector Classifier (Linear Kernel)

## Motivation

The linear SVM is the natural **baseline** before introducing more flexible kernels. It extends the hard-margin classifier by allowing some observations to fall inside the margin or even be misclassified, which makes it suitable for non-separable data such as ours.

In practice, the cost parameter $C$ controls the trade-off between margin width and classification errors. Small values of $C$ allow a softer margin and stronger regularisation, whereas large values penalise violations more heavily and tend to produce a tighter, more data-following boundary.

The linear SVM is also a natural point of comparison with Logistic Regression and LDA from Midterm 1, since all three define linear decision boundaries. This makes it a useful benchmark before moving to kernel-based SVMs, where non-linear structure can be captured more flexibly.

## Hyperparameter tuning

```{r svc-tune}
# ── Cross-validation control ────────────────────────────────────────────────
ctrl_svm <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Grid for C
grid_linear <- expand.grid(C = c(0.001, 0.01, 0.1, 1, 2, 5))

set.seed(42)

# Note: The model training is saved to an RDS object for reproducible rendering
# svc_model <- train(
#   exited ~ .,
#   data       = train_data,
#   method     = "svmLinear",
#   trControl  = ctrl_svm,
#   metric     = "ROC",
#   tuneGrid   = grid_linear,
#   preProcess = c("center", "scale")
# )

svc_model <- readRDS("models/svc_model.rds")
svc_model
```

```{r svc-tune-plot}
 svc_cv <- svc_model$results %>%
   select(C, ROC)
 
 ggplot(svc_cv, aes(x = C, y = ROC)) +
   geom_line(color = "steelblue", linewidth = 1) +
   geom_point(color = "steelblue", size = 2.2) +
  scale_x_log10() +
  labs(
    title    = "Linear SVM: Cross-Validation ROC vs C",
    x        = "Cost parameter C (log scale)",
    y        = "Cross-validated AUC (ROC)"
  ) +
  theme_minimal(base_size = 13)
```

We applied 5-fold cross-validation on the training set, and the results show that the linear SVM achieves only moderate discrimination, with the best performance at $C = 2$. The improvement over nearby values is small, which suggests that the overall limitation is not the exact choice of $C$, but the restricted flexibility of a linear boundary for this dataset. 

## Performance on the test set

```{r svc-eval, echo=FALSE}
# Obtain predicted probabilities on the test set
svc_probs <- predict(svc_model, test_data, type = "prob")[, "Yes"]

# Threshold sweep
svc_sweep <- sweep_threshold(test_data$exited, svc_probs)

# Evaluation at key thresholds
svc_results <- bind_rows(
  evaluate_model(test_data$exited, svc_probs, "SVC Linear (Cost-Sens 0.09)", threshold = pi_star)
)

knitr::kable(svc_results, caption = "SVC (Linear)")
```

```{r svc-sweep-plot, echo=FALSE}
# Optimal threshold
opt_th_svc <- svc_sweep$threshold[which.min(svc_sweep$exp_cost)]
opt_cost_svc <- min(svc_sweep$exp_cost)

svc_sweep_long <- svc_sweep %>%
  pivot_longer(c(sensitivity, specificity, exp_cost),
               names_to = "metric", values_to = "value")

ggplot(svc_sweep_long, aes(threshold, value, colour = metric)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = opt_th_svc, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = pi_star,    linetype = "dotted", colour = "grey40") +
  annotate("text", x = opt_th_svc + 0.02, y = max(svc_sweep$exp_cost) * 0.9,
           label = paste0("Optimal ≈ ", round(opt_th_svc, 2)),
           hjust = 0, size = 3.5) +
  scale_colour_viridis_d(option = "D", name = "Metric",
                         labels = c("Expected Cost", "Sensitivity", "Specificity")) +
  labs(
    title    = "SVC (Linear): Threshold Optimisation",
    subtitle = paste0("Min expected cost = ", round(opt_cost_svc, 3),
                      " at threshold ≈ ", round(opt_th_svc, 2)),
    x = "Probability Threshold", y = "Value"
  ) +
  theme_minimal(base_size = 13)
```

### Interpretation

The linear SVC provides a useful baseline, but its overall performance is only moderate. Its test AUC is 0.736, which confirms that a linear decision boundary captures part of the churn structure, but not all of it. This is consistent with Midterm 1, where the linear probabilistic models also showed limited discrimination relative to more flexible methods.

From a practical point of view, using the cost-sensitive threshold clearly improves the expected cost, but this does not change the main conclusion: the linear SVM remains constrained by the strong class overlap and the limited linear separability of the data. Its role in this project is therefore mainly as a benchmark against which more flexible kernel-based SVMs can be assessed.


# Support Vector Machine (Radial Kernel)

## Motivation

While the linear SVM is useful as a baseline, the RBF kernel allows the decision boundary to adapt more flexibly to the structure of the data, which is especially relevant in a problem with substantial class overlap and evidence of non-linearity.

We focus on the **Radial Basis Function (RBF) kernel**, which is the most standard and flexible choice in this setting. Its behaviour is mainly determined by two hyperparameters:

- **\(C\)**, which controls how strongly margin violations are penalised,
- **\(\gamma\)**, which controls how local or smooth the resulting decision boundary is.

Together, these parameters determine the balance between flexibility and overfitting.

## Hyperparameter tuning

Again, we perform a grid search with 5-fold cross-validation, optimising AUC (ROC).

```{r svm-rbf-tune, echo=TRUE, include=FALSE}
grid_rbf <- expand.grid(
  C     = c(0.1, 0.5, 1, 2, 5),       # Capped to avoid collapse
  sigma = c(0.005, 0.01, 0.1)     
)

ctrl_svm$verboseIter <- TRUE
set.seed(42)

# Note: The model training is saved to an RDS object for reproducible rendering
# svm_rbf_model <- train(
#   exited ~ .,
#   data       = train_data,
#   method     = "svmRadial",
#   trControl  = ctrl_svm,
#   metric     = "ROC",
#   tuneGrid   = grid_rbf,
#   preProcess = c("center", "scale"),
#   verboseIter     = TRUE
# )

svm_rbf_model <- readRDS("models/svm_rbf_model.rds")
```

```{r svm-rbf-heatmap, echo=FALSE}
perf_rbf <- svm_rbf_model$results %>%
  select(C, sigma, ROC)

ggplot(perf_rbf, aes(x = factor(sigma), y = factor(C), fill = ROC)) +
  geom_tile() +
  geom_text(aes(label = round(ROC, 3)), colour = "black", size = 2.8) +
  scale_fill_viridis(option = "plasma", direction = 1,
                     limits = c(min(perf_rbf$ROC), max(perf_rbf$ROC))) +
  labs(
    title    = "SVM (RBF): Grid Search — Cross-Validated AUC",
    subtitle = "5-fold CV on original training data",
    x        = expression(sigma ~ "(controls" ~ gamma ~ ")"),
    y        = "Cost C",
    fill     = "CV AUC"
  ) +
  theme_minimal(base_size = 12)
```

The best cross-validated performance was obtained at $C = 5$ and $\gamma = 0.005$, although the combination $C = 2$ and $\gamma = 0.01$ achieved virtually the same AUC. This suggests that the RBF SVM performs well over a small region of the hyperparameter space, rather than at a single sharply defined optimum. Since our goal is to compare model classes rather than to carry out exhaustive hyperparameter optimisation, we did not expand the grid further. 

Also, notice that performance drop in the top-right corner (high $C$, high $\sigma$) which illustrates overfitting, as the model constructs highly localized boundaries that capture training noise rather than generalizable patterns.

To provide an interpretable view of the fitted RBF SVM, we visualise its predicted churn probabilities in the Age–Products plane, fixing the remaining variables at a representative customer profile. Among the candidate views we explored, this was the most interpretable one. Even so, it should only be read as a partial illustration of the model, since the classifier is fitted in a much higher-dimensional space and the full decision rule cannot be faithfully represented in two dimensions. This is why no plot was shown in SVM Soft Margin, no 2D representation was able to capture the structure. 

```{r svm-profile-setup, include=FALSE}
# ── Reference profile for visualizations ────────────────────────────────
# Ensure card_type is a factor
train_data <- train_data %>%
  mutate(card_type = factor(card_type))

# Medians for continuous variables
medians <- train_data %>%
  summarise(
    credit_score       = median(credit_score),
    tenure             = median(tenure),
    estimated_salary   = median(estimated_salary),
    satisfaction_score = median(satisfaction_score),
    point_earned       = median(point_earned),
    num_of_products    = as.integer(median(num_of_products))
  )

# Modes for categorical variables
modal <- list(
  geography        = names(sort(table(train_data$geography),        decreasing = TRUE))[1],
  gender           = names(sort(table(train_data$gender),           decreasing = TRUE))[1],
  has_cr_card      = names(sort(table(train_data$has_cr_card),      decreasing = TRUE))[1],
  is_active_member = names(sort(table(train_data$is_active_member), decreasing = TRUE))[1],
  card_type        = names(sort(table(train_data$card_type),        decreasing = TRUE))[1]
)

# Sequences for the grids
age_seq     <- seq(min(train_data$age),     max(train_data$age),     length = 150)
balance_seq <- seq(min(train_data$balance), max(train_data$balance), length = 150)
age_seq2    <- seq(min(train_data$age),     max(train_data$age),     length = 200)
```

```{r svm-rbf-grid-plot, echo=FALSE}
# ── Grid: age × num_of_products (faceted) ──────────────────────────────────
age_seq2 <- seq(min(train_data$age), max(train_data$age), length = 200)

grid_prod <- expand.grid(
  age             = age_seq2,
  num_of_products = c(1, 2, 3, 4)
) %>%
  mutate(
    credit_score       = medians$credit_score,
    tenure             = medians$tenure,
    estimated_salary   = medians$estimated_salary,
    satisfaction_score = medians$satisfaction_score,
    point_earned       = medians$point_earned,
    balance            = median(train_data$balance),
    geography        = factor("Germany", levels = levels(train_data$geography)),
    gender           = factor("Male",  levels = levels(train_data$gender)),
    has_cr_card      = factor("Yes",     levels = levels(train_data$has_cr_card)),
    is_active_member = factor("No",      levels = levels(train_data$is_active_member)),
    card_type        = factor(modal$card_type, levels = levels(train_data$card_type))
  )

grid_prod$pred  <- predict(svm_rbf_model, grid_prod)
grid_prod$prob  <- predict(svm_rbf_model, grid_prod, type = "prob")[, "Yes"]

# Real data points
set.seed(42)
train_plot2 <- train_data %>% slice_sample(n = 2000)

ggplot() +
  # Continuous probability background
  geom_tile(data = grid_prod,
            aes(age, num_of_products, fill = prob), alpha = 0.65) +
  # Decision boundary (prob = 0.5 and prob = pi_star)
  geom_contour(data = grid_prod,
               aes(age, num_of_products, z = prob),
               breaks = 0.12,     colour = "black",   linewidth = 1,   linetype = "solid") +
  # Real data points
  geom_jitter(data = train_plot2,
              aes(age, num_of_products, colour = exited),
              width = 0, height = 0.15, size = 0.9, alpha = 0.85) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#d6604d",
    midpoint = 0.5,
    name     = "P(Churn)"
  ) +
  scale_colour_manual(
    values = c("No" = "#2166ac", "Yes" = "#d6604d"),
    name   = "Actual class"
  ) +
  scale_y_continuous(breaks = 1:4) +
  labs(
    title    = "SVM (RBF): Churn probability — Age × Number of Products",
    subtitle = paste0("Fixed profile: Germany, Male, inactive, balance = median",
                      "\nThreshold = 0.12"),
    x = "Age", y = "Number of products"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

The plot shows that the model does not induce a purely linear separation: the decision boundary changes with age and number of products. In particular, the probability of churn increases markedly at older ages, but the effect is not uniform across product levels, suggesting an interaction structure that a linear classifier cannot capture as flexibly.

## Test-set performance

```{r svm-rbf-eval, echo=FALSE}
rbf_probs <- predict(svm_rbf_model, test_data, type = "prob")[, "Yes"]

rbf_sweep <- sweep_threshold(test_data$exited, rbf_probs)

rbf_results <- bind_rows(
  evaluate_model(test_data$exited, rbf_probs, "SVM RBF (Cost-Sens 0.09)", threshold = pi_star)
)

knitr::kable(rbf_results, caption = "SVM (RBF Kernel)")
```

```{r svm-rbf-sweep-plot, echo=FALSE}
opt_th_rbf   <- rbf_sweep$threshold[which.min(rbf_sweep$exp_cost)]
opt_cost_rbf <- min(rbf_sweep$exp_cost)

rbf_sweep_long <- rbf_sweep %>%
  pivot_longer(c(sensitivity, specificity, exp_cost),
               names_to = "metric", values_to = "value")

ggplot(rbf_sweep_long, aes(threshold, value, colour = metric)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = opt_th_rbf, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = pi_star,    linetype = "dotted", colour = "grey40") +
  annotate("text", x = opt_th_rbf + 0.02, y = max(rbf_sweep$exp_cost) * 0.9,
           label = paste0("Optimal ≈ ", round(opt_th_rbf, 2)),
           hjust = 0, size = 3.5) +
  scale_colour_viridis_d(option = "D", name = "Metric",
                         labels = c("Expected Cost", "Sensitivity", "Specificity")) +
  labs(
    title    = "SVM (RBF): Threshold Optimisation",
    subtitle = paste0("Min expected cost = ", round(opt_cost_rbf, 3),
                      " at threshold ≈ ", round(opt_th_rbf, 2)),
    x = "Probability Threshold", y = "Value"
  ) +
  theme_minimal(base_size = 13)
```

We can see that the empirical minimum of the cost curve is reached near 0.12, which is close to the theoretical cost-sensitive threshold and supports the use of $\pi^* = 0.09$ as a reasonable operating rule.

```{r svm-roc, echo=FALSE}
roc_svc <- roc(test_data$exited, svc_probs,  levels = c("No","Yes"), quiet = TRUE)
roc_rbf <- roc(test_data$exited, rbf_probs,  levels = c("No","Yes"), quiet = TRUE)

ggroc(list(
  "SVC Linear" = roc_svc,
  "SVM RBF"    = roc_rbf
), linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey50") +
  scale_colour_brewer(palette = "Set1", name = "Model") +
  annotate("text", x = 0.4, y = 0.55,
           label = paste0("SVC AUC = ", round(auc(roc_svc), 3))) +
  annotate("text", x = 0.4, y = 0.48,
           label = paste0("RBF AUC = ", round(auc(roc_rbf), 3))) +
  labs(
    title    = "ROC Curves: SVC vs SVM (RBF) on Test Set",
    subtitle = "Threshold-independent comparison of ranking performance"
  ) +
  theme_minimal(base_size = 13)
```

The RBF SVM clearly outperforms the linear SVM. Its test AUC rises, which indicates that allowing for a non-linear boundary captures the churn structure much better than a linear classifier.

This improvement is also reflected in the expected cost: at the cost-sensitive threshold, the RBF SVM achieves 6.463, compared with 7.184 for the linear SVM, while maintaining very high sensitivity (0.929).  This makes it the stronger SVM model for our problem.

Overall, these results confirm the main evidence from Midterm 1: non-linear methods are better suited to this dataset than purely linear classifiers.


# Support Vectors and Class Overlap 

A key property of SVMs is that the fitted boundary depends only on a subset of the training observations: the **support vectors**. These are the points that lie on, or close to, the margin and therefore determine the position of the classifier. Observations far from the boundary do not influence the final decision rule.

This is one of the main differences with LDA and QDA, whose fitted boundaries depend on summary statistics computed from all training observations. For this reason, the number of support vectors can give useful information about how difficult the classification problem is.

We found substantial overlap in our dataset between churners and non-churners. This has a direct implication for SVMs, when the classes are heavily mixed, more observations tend to lie close to the boundary or inside the margin, so a larger fraction of the training set becomes support vectors.

In this sense, the proportion of support vectors provides a simple quantitative summary of the overlap already observed in Midterm 1. A high proportion would be consistent with a difficult classification problem and limited class separation.

## Number of support vectors per model

```{r sv-extraction, include=FALSE}
# Both models use kernlab (ksvm) → same method to extract SVs
svc_raw <- svc_model$finalModel
rbf_raw <- svm_rbf_model$finalModel

svc_sv_total <- kernlab::nSV(svc_raw)
rbf_sv_total <- kernlab::nSV(rbf_raw)

sv_summary <- tibble(
  Model    = c("SVC (Linear)", "SVM (RBF)"),
  SV_Total = c(svc_sv_total, rbf_sv_total),
  N_Train  = nrow(train_data)
) %>%
  mutate(Pct_SV = round(100 * SV_Total / N_Train, 1))
```

```{r sv-barplot, echo=FALSE}
ggplot(sv_summary, aes(Model, Pct_SV, fill = Model)) +
  geom_col(width = 0.4, alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0(Pct_SV, "%\n(n = ", SV_Total, ")")),
            vjust = -0.3, size = 4.2) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    title    = "Proportion of training observations acting as support vectors",
    subtitle = paste0("Training set size = ", nrow(train_data)),
    x = "", y = "% of training observations"
  ) +
  theme_minimal(base_size = 13)
```

```{r sv-interpretation, echo=FALSE}
pct_svc <- sv_summary$Pct_SV[sv_summary$Model == "SVC (Linear)"]
pct_rbf <- sv_summary$Pct_SV[sv_summary$Model == "SVM (RBF)"]
n_svc   <- sv_summary$SV_Total[sv_summary$Model == "SVC (Linear)"]
n_rbf   <- sv_summary$SV_Total[sv_summary$Model == "SVM (RBF)"]
```

We can see that the linear SVC uses **`r pct_svc`%** of the training set as support vectors, whereas the RBF SVM uses **`r pct_rbf`%**. Both proportions are relatively high, which is consistent with the substantial class overlap already predicted.

The difference between the two models is also informative. The RBF SVM achieves better predictive performance while relying on a smaller proportion of support vectors. This suggests that, once non-linearity is allowed, the classes can be separated more efficiently in the transformed feature space than under a purely linear boundary.

```{r knn-split-up, echo = FALSE}
set.seed(42)

# Imbalanced Strategy (just oversampling)
# Train / Test Split (80/20) 
trainIndex <- createDataPartition(df$exited, p = 0.8, list = FALSE)
train_imb <- df[trainIndex, ]
test_imb  <- df[-trainIndex, ]

# Oversampling
train_x <- train_imb %>% dplyr::select(-exited)
train_y <- train_imb$exited
train_up <- upSample(x = train_x, y = train_y, yname = "exited")
```

# K-Nearest Neighbors (K-NN)

To gain an initial understanding of the KNN classifier, we first focus on only two numerical predictors, Age and Balance, so that the decision regions can be visualized. Since KNN is based on distances, these variables must be scaled.

In this case, to handle class imbalance, we will only use oversampling, following the same approach adopted in the first part of the project. Notice, that the best results, in terms of expected cost, were obtained with the theoretical threshold, $t^{*} = 0.09$, which will therefore be maintained in this analysis.

## K-NN with only two predictors

```{r knn-prep-2d, echo = FALSE}
# Dataset we are going to use: train = train_up, test = test_imb

num_cols <- names(train_up)[sapply(train_up, is.numeric)]

# Separate X and Y
x_train_up <- subset(train_up, select = -exited)
x_test_imb  <- subset(test_imb, select = -exited)

y_train_up <- train_up$exited
y_test_imb  <- test_imb$exited

# Copies for scaling
x_train_proc <- x_train_up
x_test_proc  <- x_test_imb

# Fit scaling ONLY on the training set
pp <- preProcess(x_train_up[, num_cols], method = c("center", "scale"))

# Apply to numerical columns
x_train_proc[, num_cols] <- predict(pp, x_train_up[, num_cols])
x_test_proc[, num_cols]  <- predict(pp, x_test_imb[, num_cols])
```

```{r knn-grid-2d, echo = FALSE}
# ------------------------------------------------------------------
# 1) TEST points
# ------------------------------------------------------------------
plot_test <- data.frame(
  age = x_test_proc$age,
  balance = x_test_proc$balance,
  exited = y_test_imb
)

# ------------------------------------------------------------------
# 2) Grid for decision boundaries
#    (boundary is built using TRAIN data)
# ------------------------------------------------------------------
x_range <- seq(
  min(x_train_proc$age) - 0.5,
  max(x_train_proc$age) + 0.5,
  length.out = 200
)

y_range <- seq(
  min(x_train_proc$balance) - 0.5,
  max(x_train_proc$balance) + 0.5,
  length.out = 200
)

grid <- expand.grid(
  age = x_range,
  balance = y_range
)

# Prediction on the grid using TRAIN
grid$k1 <- knn(
  train = x_train_proc[, c("age", "balance")],
  test  = grid[, c("age", "balance")],
  cl    = y_train_up,
  k     = 1
)

grid$k5 <- knn(
  train = x_train_proc[, c("age", "balance")],
  test  = grid[, c("age", "balance")],
  cl    = y_train_up,
  k     = 5
)

grid$k10 <- knn(
  train = x_train_proc[, c("age", "balance")],
  test  = grid[, c("age", "balance")],
  cl    = y_train_up,
  k     = 10
)

# ------------------------------------------------------------------
# 3) Plotting function
# ------------------------------------------------------------------

make_knn_plot_grid <- function(grid_col, k_value) {
  ggplot() +
    geom_tile(
      data = grid,
      aes(x = age, y = balance, fill = .data[[grid_col]]),
      alpha = 0.25
    ) +
    scale_fill_manual(values = my_colors,
                      labels = c("No" = "Non-churn", "Yes" = "Churn")) +
    labs(
      title = paste0("KNN with k = ", k_value),
      x = "Age (scaled)",
      y = "Balance (scaled)",
      fill = "Predicted region"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = 8),
      plot.title = element_text(size = 9, face = "bold",hjust = 0.5)
    )
}

make_knn_plot <- function(grid_col, k_value) {
  ggplot() +
    geom_tile(
      data = grid,
      aes(x = age, y = balance, fill = .data[[grid_col]]),
      alpha = 0.25
    ) +
    geom_point(
      data = plot_test,
      aes(x = age, y = balance, color = exited),
      size = 0.5,
      alpha = 1
    ) +
    scale_fill_manual(values = my_colors,
                      labels = c("No" = "Non-churn", "Yes" = "Churn")) +
    scale_color_manual(values = my_colors,  labels = c("No" = "Non-churn", "Yes" = "Churn")) +
    labs(
      title = paste0("KNN with k = ", k_value),
      x = "Age (scaled)",
      y = "Balance (scaled)",
      color = "Observed class",
      fill = "Predicted region"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = 8),
      plot.title = element_text(size = 9, face = "bold", hjust = 0.5))
}

# ------------------------------------------------------------------
# 4) Charts for k = 1, 5, 10
# ------------------------------------------------------------------
p_k1  <- make_knn_plot_grid("k1", 1)+ theme(legend.position = "none")
pp_k1  <- make_knn_plot("k1", 1)+ theme(legend.position = "none")

p_k5  <- make_knn_plot_grid("k5", 5) + theme(legend.position = "bottom")
pp_k5  <- make_knn_plot("k5", 5) + theme(legend.position = "bottom")

p_k10 <- make_knn_plot_grid("k10", 10) +theme(legend.position = "none")
pp_k10 <- make_knn_plot("k10", 10) +theme(legend.position = "none")
```

```{r knn-grid-plot, fig.width=10, echo = FALSE}
(p_k1 | p_k5 | p_k10) +  plot_annotation(title = "KNN Decision Boundaries: Effect of k",
                  subtitle = "Small k = complex boundary (overfitting), Large k = smooth boundary (underfitting)")
```

The main point is the following: as $k$ increases, the decision boundary becomes smoother and more stable. For $k = 1$, the predicted regions, in particular the blue one, contain many small irregularities and abrupt class changes (high variance). By contrast, larger values of $k$ produce somehow more regular regions. This pattern is consistent with the bias-variance trade-off.

The figure also suggests that the **classification boundary is influenced more strongly by Age than by Balance**, since most of the variation in the predicted regions occurs along the horizontal axis. In this two-dimensional representation, Age appears to drive the main separation between classes, while Balance plays a more secondary role, mainly adjusting the boundary in specific areas of the predictor space.

Notice that the **churn class is concentrated mainly in the central area of the predictor space**, while the Non-churn region dominates more at the extremes. This reveals that the relationship between these variables and churn does not seem to be simply linear, which helps explain why KNN could be appropriate in this setting.

```{r knn-test-plot, fig.width=10, echo = FALSE}
(pp_k1 | pp_k5 | pp_k10) + plot_annotation(title = "k-NN Decision Boundaries: Effect of k",
                                           subtitle = "Test set observations were added")
```

Finally, when the test set observations are added to the previous plot, it becomes clear that, **when KNN is fitted using only Age and Balance, the two classes still show substantial overlap**. This suggests that these two predictors alone do not provide a sufficiently clear separation between churned and non-churned customers.

For this reason, we next extend the model by including all available variables and fit a new KNN model. The optimal value of $k$ is then selected by cross-validation, according to our business objective, minimizing the expected cost.

## K-NN full model

```{r knn-cost-setup, echo = FALSE}
# =========================================================
# 2) Costs and threshold
# =========================================================
threshold_cost <- 0.09
cost_fn <- 100
cost_fp <- 10
positive_class <- "Yes"

# =========================================================
# 3) Custom metric: Expected Cost: computed from yes 
#    probabilities
# =========================================================
costSummary <- function(data, lev = NULL, model = NULL,
                        threshold = 0.09,
                        cost_fn = 100,
                        cost_fp = 10,
                        positive_class = "Yes") {
  
  negative_class <- setdiff(lev, positive_class)
  
  probs <- data[[positive_class]]
  
  preds <- factor(
    ifelse(probs >= threshold, positive_class, negative_class),
    levels = lev
  )
  
  obs <- factor(data$obs, levels = lev)
  
  fn <- sum(preds == negative_class & obs == positive_class)
  fp <- sum(preds == positive_class & obs == negative_class)
  
  exp_cost <- (fn * cost_fn + fp * cost_fp) / length(obs)
  
  c(ExpectedCost = exp_cost)
}
```

```{r knn-cv}
# Set cross-validation 
ctrl_cost <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = function(data, lev = NULL, model = NULL) {
    costSummary(
      data = data,
      lev = lev,
      model = model,
      threshold = threshold_cost,
      cost_fn = cost_fn,
      cost_fp = cost_fp,
      positive_class = positive_class
    )
  },
  sampling = "up",
  savePredictions = "final"
)

# Fit the model using our expected cost criterion
set.seed(42)

knn_cv_cost <- train(
  exited ~ .,
  data = train_imb,
  method = "knn",
  trControl = ctrl_cost,
  preProcess = c("center", "scale"),
  tuneGrid = data.frame(k = 1:20),
  metric = "ExpectedCost",
  maximize = FALSE
)
```

```{r knn-cv-plot, echo = FALSE, fig.width=10}
# =========================================================
# 6) Expected Cost plot by k
# =========================================================
ggplot(knn_cv_cost$results, aes(x = k, y = ExpectedCost)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = knn_cv_cost$bestTune$k,
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Cross-validation (K=5) using Expected Cost for k-NN",
    x = "k",
    y = "Expected Cost"
  ) +
  theme_minimal()
```

The cross-validation expected cost curve shows a clear pattern: **For very small values of k, the expected cost is relatively high**, which suggests that **the classifier is overly sensitive to noise** and therefore suffers from **high variance**. As **k increases, the expected cost decreases significantly**, showing that growing the neighborhood substantially improves stability. 

After the initial decline, the curve becomes much flatter, and **from around $k=7$ onward the model enters a more stable region**. Although the minimum expected cost is achieved at $k=13$, the additional reduction beyond this point is negligible and does not justify the increase in model complexity.

```{r knn-test-eval, echo =FALSE}
# =========================================================
# 7) Test evaluation function
# =========================================================
evaluate_model_ <- function(model_fit, test_data, true_labels,
                           threshold = 0.09,
                           cost_fn = 100,
                           cost_fp = 10,
                           positive_class = "Yes") {
  
  true_labels <- factor(true_labels, levels = c("No", "Yes"))
  
  probs <- predict(model_fit, newdata = test_data, type = "prob")[, positive_class]
  
  preds <- factor(
    ifelse(probs >= threshold, positive_class, "No"),
    levels = c("No", "Yes")
  )
  
  cm <- confusionMatrix(preds, true_labels, positive = positive_class)
  
  fn <- sum(preds == "No"  & true_labels == positive_class)
  fp <- sum(preds == positive_class & true_labels == "No")
  exp_cost <- (fn * cost_fn + fp * cost_fp) / length(true_labels)
  
  roc_obj <- roc(response = true_labels, predictor = probs, quiet = TRUE)
  
  tibble(
    AUC               = as.numeric(auc(roc_obj)),
    Accuracy          = unname(cm$overall["Accuracy"]),
    Sensitivity       = unname(cm$byClass["Sensitivity"]),
    Specificity       = unname(cm$byClass["Specificity"]),
    Precision         = unname(cm$byClass["Precision"]),
    F1_Score          = unname(cm$byClass["F1"]),
    Balanced_Accuracy = unname(cm$byClass["Balanced Accuracy"]),
    Expected_Cost     = exp_cost
  )
}

# =========================================================
# 8) Final test evaluation
# =========================================================
knn_test_eval <- evaluate_model_(
  model_fit = knn_cv_cost,
  test_data = test_imb,
  true_labels = test_imb$exited,
  threshold = threshold_cost,
  cost_fn = cost_fn,
  cost_fp = cost_fp,
  positive_class = positive_class
)

# =========================================================
# 9) Predictions and probabilities on test
# =========================================================
prob_test <- predict(knn_cv_cost, newdata = test_imb, type = "prob")[, "Yes"]

pred_test <- factor(
  ifelse(prob_test >= threshold_cost, "Yes", "No"),
  levels = c("No", "Yes")
)

cm_test <- confusionMatrix(pred_test, test_imb$exited, positive = "Yes")

# =========================================================
# 10) Selected k
# =========================================================
k_opt <- knn_cv_cost$bestTune$k
```

Based on the observed pattern, **we select $k=6$ as the final choice**. In particular, it is large enough to avoid the instability associated with very small neighbourhoods, while still being small enough to retain a relatively local and adaptive decision boundary. Moreover, it represents an appropriate choice for our business application.

```{r knn-final}
# Scale features on train
X_train_up <- model.matrix(exited ~ . - 1, data = train_up) |> as.data.frame()
X_test     <- model.matrix(exited ~ . - 1, data = test_imb) |> as.data.frame()

pp <- preProcess(X_train_up, method = c("center", "scale"))

X_train_scaled <- predict(pp, X_train_up)
X_test_scaled  <- predict(pp, X_test)

# Fit the final knn model
knn_pred_exp <- knn(
  train = X_train_scaled,
  test  = X_test_scaled,
  cl    = train_up$exited,
  k     = 6,
  prob  = TRUE
)

# Evaluate the model
knn_eval <- evaluate_knn(
  pred_knn = knn_pred_exp,
  true_labels = test_imb$exited,
  threshold = thresh_theo
)
```

```{r knn-final-table, echo = FALSE}
knitr::kable(knn_eval, digits = 3, 
             caption = "KNN: Optimal model")
```

# Tree Based Models

## Decision Tree

We now turn to decision trees and follow a structure similar to the one used for KNN. We begin with a simplified setting based on only two predictors, Age and Balance, in order to visualize how the tree partitions the predictor space. We then extend the analysis by fitting a new tree using the full set of predictors.

```{r tree-2d, echo = FALSE}
set.seed(42)

churn_tree_2d <- tree(
  exited ~ age + balance,
  data = train_up,
  control = tree.control(
    nobs    = nrow(train_up),
    mindev  = 0.01,
    minsize = 5
  )
)

#-----------------------------
# Create grid for partition representation
#-----------------------------
churn_grid <- expand.grid(
  age = seq(min(train_up$age) - 1,
            max(train_up$age) + 1,
            length.out = 200),
  balance = seq(min(train_up$balance) - 100,
                max(train_up$balance) + 100,
                length.out = 200)
)

# Predict class for each point on the grid
churn_grid$Pred <- predict(churn_tree_2d, newdata = churn_grid, type = "class")
```

The **tree partition is driven almost entirely by Age**, as shown by the vertical decision regions. This suggests that the model relies mainly on age thresholds, while Balance contributes little to the final partition. A similar effect was seen when using KNN. In particular, the tree identifies an intermediate age range as the region with the highest likelihood of churn (orange), whereas lower and higher ages are mostly classified as non-churn (blue).

```{r tree-2d-plot, echo = FALSE}
#-----------------------------
# Space Partition Plot
#-----------------------------
ggplot() +
  geom_tile(
    data = churn_grid,
    aes(x = age, y = balance, fill = Pred),
    alpha = 0.28
  ) +
  geom_point(
    data = test_imb,
    aes(x = age, y = balance, color = exited),
    size = 1.1,
    alpha = 0.85
  ) +
  labs(
    title = "Decision tree partition in the Age–Balance feature space",
    subtitle = "Background: class predicted by the tree",
    x = "Age",
    y = "Balance"
  ) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) +
  theme_minimal()
```

Although this partition captures some general structure in the data, **there is still substantial overlap between the two classes** within each region, especially in the central band predicted as Yes. Therefore, the tree yields a simple and clear classification rule. However, Age and Balance alone do not seem sufficient to achieve a clear separation between churned and non-churned customers.

To obtain the final tree model, we first grew a large initial tree and then pruned it using cross-validation
using expected cost rather than misclassification error. This criterion is more appropriate in our churn setting, since the data are imbalanced and false negatives are more costly than false positives.

```{r tree-cv}
# Matrix cost
loss_mat <- matrix(
  c(0,   10,
    100, 0),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    actual    = c("No", "Yes"),
    predicted = c("No", "Yes")
  )
)

# Fit full tree
set.seed(42)
churn_big <- tree(
  exited ~ .,
  data = train_imb,
  control = tree.control(
    nobs = nrow(train_imb),
    mindev = 0,
    minsize = 2
  )
)

## Cross validation for pruning
set.seed(42)

churn_cv <- cv.tree(
  churn_big,
  FUN = prune.misclass,
  K = 10,
  loss = loss_mat
)
```

```{r tree-prune-plot, echo = FALSE, fig.width=10}
par(mfrow = c(1, 2))
plot(
  churn_cv$size,
  churn_cv$dev / nrow(train_imb),
  type = "b",
  pch = 19,
  xlab = "Tree size",
  ylab = "Cross-validated expected cost",
  main = "Cost-sensitive CV for tree pruning",
  xlim = c(0, 10)
)

best_size <- 4
  
churn_pruned <- prune.misclass(
  churn_big,
  best = best_size,
  loss = loss_mat
)

plot(churn_pruned)
text(churn_pruned, pretty = 0, cex = 0.85)
title(paste("Pruned tree ", best_size, "leaves"))

par(mfrow = c(1, 1))
```

After applying cross-validation, the minimum expected cost was attained at size = 572. However, such a large tree is unlikely to generalize well to new observations, as it reflects an excessively complex model. If we zoom into the plot and focus on the first tree sizes, we see that the expected cost drops sharply at the beginning, but from around size = 5 onward the improvement becomes negligible. In other words, increasing complexity beyond that point provides little practical benefit.

For this reason, we selected a tree size of 4, meaning that the final pruned tree contains four terminal nodes. This suggests that a relatively simple tree is enough to capture the main structure of the classification problem while still performing well under the cost-sensitive validation criterion.

This pruned tree highlights a simple but intuitive churn pattern: **age is the primary split, with product holding and member activity determining the final decision**.

```{r tree-eval, echo = FALSE}
# =========================================================
# Probabilistic predictions on test
# For classification, type = "vector" returns class probabilities
# =========================================================
prob_tree <- predict(churn_pruned, newdata = test_imb, type = "vector")[, "Yes"]

# =========================================================
# Convert probabilities to classes
# =========================================================
thresh_theo <- 0.1
pred_tree <- factor(
  ifelse(prob_tree >= thresh_theo, "Yes", "No"),
  levels = c("No", "Yes")
)

# =========================================================
# Final test evaluation
# =========================================================
cm_tree <- confusionMatrix(pred_tree, test_imb$exited, positive = "Yes")

# =========================================================
# Expected Cost on test
# =========================================================
fn <- sum(pred_tree == "No"  & test_imb$exited == "Yes")
fp <- sum(pred_tree == "Yes" & test_imb$exited == "No")

exp_cost_tree <- (fn * 100 + fp * 10) / nrow(test_imb)

# =========================================================
# AUC on test
# =========================================================
roc_tree <- roc(response = test_imb$exited, predictor = prob_tree, quiet = TRUE)
auc_tree <- as.numeric(auc(roc_tree))

# =========================================================
# Final summary table
# =========================================================
tree_results <- tibble(
  Best_Tree_Size     = best_size,
  Threshold_Used     = threshold_cost,
  AUC                = auc_tree,
  Accuracy           = unname(cm_tree$overall["Accuracy"]),
  Sensitivity        = unname(cm_tree$byClass["Sensitivity"]),
  Specificity        = unname(cm_tree$byClass["Specificity"]),
  Precision          = unname(cm_tree$byClass["Precision"]),
  F1_Score           = unname(cm_tree$byClass["F1"]),
  Balanced_Accuracy  = unname(cm_tree$byClass["Balanced Accuracy"]),
  Expected_Cost      = exp_cost_tree
)
```

```{r tree-table, echo = FALSE}
knitr::kable(
  dplyr::select(tree_results, -Best_Tree_Size, -Threshold_Used),
  digits = 3,
  caption = "Decision Tree: Optimal model"
)
```

## Random Forest

### Bagging 

Although decision trees are easy to interpret, they are often unstable and sensitive to small changes in the data. To overcome these limitations and improve predictive performance, we now move to Random Forests, which combine many trees built on bootstrap samples and aggregate their predictions.

We start with bagging, a special case of Random Forest where all predictors are considered at each split:

```{r bagging}
# Number of predictors
p_cls <- ncol(train_up) - 1
thresh_theo <- 0.09

# Set number of trees values for ensembles and Cross Validation folds
ntree_fast <- 250
ntree_main <- 350
cv_folds <- 5

set.seed(42)

bag_cls <- randomForest(
  exited ~ ., data = train_up,
  mtry = p_cls,
  ntree = ntree_main,
  importance = TRUE
)

bagging_up <- evaluate_model_(model_fit = bag_cls, test_data = test_imb, true_labels = test_imb$exited, threshold = thresh_theo)
```

```{r bagging-table, echo = FALSE}
knitr::kable(bagging_up, digits = 3, 
             caption = "Bagging: Optimal model")
```

### Tuning number of predictors (mtry)

Next we tune mtry with OOB error and fit a random forest.

```{r rf-mtry-tuning}
base_mtry_cls <- max(1, floor(sqrt(p_cls)))

# We test with different number of predictors, +1-1 and all (bagging)
mtry_grid_cls <- sort(unique(pmax(1, pmin(p_cls, c(base_mtry_cls - 1, base_mtry_cls, base_mtry_cls + 1, p_cls)))))

rf_cls_tuning <- map_dfr(mtry_grid_cls, function(m) {
  set.seed(42 + m)
  fit <- randomForest(exited ~ ., data = train_up, mtry = m, ntree = ntree_fast)

  tibble(
    mtry = m,
    OOB_Error = fit$err.rate[ntree_fast, "OOB"]
  )
})

best_mtry_cls <- rf_cls_tuning$mtry[which.min(rf_cls_tuning$OOB_Error)]

set.seed(42)
rf_cls <- randomForest(
  exited ~ ., data = train_up,
  mtry = best_mtry_cls,
  ntree = ntree_main,
  importance = TRUE
)

rf_up <- evaluate_model_(model_fit = rf_cls, test_data = test_imb, true_labels = test_imb$exited, threshold = thresh_theo)
```

```{r rf-tuning-plot, echo = FALSE, fig.width=10}
oob_ntree_df <- tibble(
  Trees = seq_len(nrow(rf_cls$err.rate)),
  OOB_Error = rf_cls$err.rate[, "OOB"]
)

p_oob_ntree <- ggplot(oob_ntree_df, aes(x = Trees, y = OOB_Error)) +
  geom_line(color = "#2c7fb8", linewidth = 1) +
  labs(title = "Churn RF: OOB Error vs Number of Trees", x = "Number of trees", y = "OOB error") +
  theme_minimal()

p_oob_mtry <- ggplot(rf_cls_tuning, aes(x = mtry, y = OOB_Error)) +
  geom_line(color = "#41ab5d", linewidth = 1) +
  geom_point(color = "#41ab5d", size = 2.4) +
  geom_vline(xintercept = best_mtry_cls, linetype = "dashed", color = "red") +
  labs(title = "Churn RF: OOB Error vs mtry", x = "mtry", y = "OOB error") +
  theme_minimal()

p_oob_ntree + p_oob_mtry
```

The **OOB error drops substantially for the first trees**, suggesting that the random forest quickly gains predictive power as the ensemble grows. After roughly 100–150 trees, the curve becomes nearly flat and stabilizes around 3%, suggesting that the model has essentially converged. Therefore, adding more trees beyond this point yields only marginal improvements, while no evidence of overfitting is observed, since the OOB error does not increase.

Among the candidate values tested for `mtry`, the lowest OOB error is achieved at $mtry = 3$. This suggests that the random forest performs best when each split considers a small, but not excessively restricted, subset of predictors. In this case, it appears to provide the best trade-off between tree diversity and split quality.

```{r rf-importance-plot, echo = FALSE, fig.width=10}
imp_cls <- as.data.frame(importance(rf_cls))
imp_cls$Variable <- rownames(imp_cls)

importance_cols <- intersect(c("MeanDecreaseAccuracy", "MeanDecreaseGini"), names(imp_cls))

imp_cls_long <- imp_cls %>%
  dplyr::select(Variable, dplyr::all_of(importance_cols)) %>%
  tidyr::pivot_longer(
    cols = -Variable,
    names_to = "Measure",
    values_to = "Importance"
  ) %>%
  group_by(Measure) %>%
  mutate(Importance_pct = 100 * Importance / sum(Importance, na.rm = TRUE)) %>%
  ungroup()

ggplot(
  imp_cls_long,
  aes(x = fct_reorder(Variable, Importance_pct), y = Importance_pct, fill = Measure)
) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "MeanDecreaseAccuracy" = "#FF7F50",
      "MeanDecreaseGini" = "#00BFFF"
    )
  ) +
  labs(
    title = "Churn RF Variable Importance (%)",
    x = "",
    y = "Relative Importance (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.78, 0.25),   # inside the plot
    legend.background = element_rect(fill = "white", colour = "grey70"),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )
```

We also inspect variable importance from the fitted random forest, the following insights were obtained:  

**Age stands out as the dominant predictor of churn**, ranking first under both decrease measures. This suggests that customer age carries the strongest signal for separating churners from non-churners in this model. A **second group of influential variables includes number of products, balance, estimated salary, credit score, and points earned**, all of which make substantial contributions to predictive performance. In contrast, variables such as *gender, has credit card, and is active member* appear to play a much smaller role, indicating that they **add relatively limited discriminatory power** once the stronger predictors are already taken into account.

The percentage representation also shows that importance is somewhat more concentrated under Mean Decrease Gini, whereas under Mean Decrease Accuracy it is distributed more evenly across several predictors. This suggests that although age is the single most influential variable, predictive performance still depends on the combined contribution of other customer and account-related features.

In conclusion, the ranking suggests that **churn in this dataset is driven more by customer profile** and **account-related characteristics** than by simple demographic or binary status indicators. Importantly, the consistency between the two importance measures strengthens this conclusion: **the variables that most improve node purity are also those whose perturbation most harms predictive accuracy**.

```{r rf-table, echo = FALSE}
knitr::kable(rf_up, digits = 3, 
             caption = "Random Forest: Optimal model")
```

## Gradient Boosting

We fit a compact GBM grid using caret.

```{r gbm-tune}
cls_ctrl <- trainControl(
  method = "cv",
  number = cv_folds,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

cls_grid <- expand.grid(
  interaction.depth = c(1, 2),
  n.trees = c(100, 250),
  shrinkage = c(0.05, 0.1),
  n.minobsinnode = c(10)
)

set.seed(42)
gbm_cls <- train(
  exited ~ ., data = train_up,
  method = "gbm",
  trControl = cls_ctrl,
  tuneGrid = cls_grid,
  metric = "ROC",
  verbose = FALSE
)
```

```{r gbm-table, echo=FALSE}
gbm_cls_up <- evaluate_model_(model_fit = gbm_cls, test_data = test_imb, true_labels = test_imb$exited, threshold = thresh_theo)

knitr::kable(gbm_cls_up, digits = 3, 
             caption = "Gradient Boosting: Optimal model")
```

## XGBoost

We also fit an XGBoost model:

```{r xgboost-tune}
xgb_terms_cls <- terms(exited ~ ., data = train_up)

# Remove intercept
x_train_xgb_cls <- model.matrix(xgb_terms_cls, data = train_up)[, -1]
x_test_xgb_cls <- model.matrix(xgb_terms_cls, data = test_imb)[, -1]
y_train_xgb_cls <- as.numeric(train_up$exited == "Yes")

set.seed(42 + 10)
xgb_dtrain_cls <- xgb.DMatrix(data = x_train_xgb_cls, label = y_train_xgb_cls)
xgb_dtest_cls <- xgb.DMatrix(data = x_test_xgb_cls)

xgb_cls <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = 0.05,
    max_depth = 3,
    subsample = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    gamma = 0
  ),
  data = xgb_dtrain_cls,
  nrounds = 250,
  verbose = 0
)
```

```{r xgboost-table, echo=FALSE}
xgb_cls_prob_yes <- as.numeric(predict(xgb_cls, newdata = xgb_dtest_cls))
xgb_cls_pred <- factor(ifelse(xgb_cls_prob_yes >= thresh_theo, "Yes", "No"), levels = c("No", "Yes"))

xgb_up <- evaluate_model_xgb(xgb_cls_pred, test_imb$exited, xgb_cls_prob_yes)

knitr::kable(xgb_up, digits = 3, 
             caption = "XGBoost: Optimal model")
```

# Conclusions 

We first show the comparison between the two principal SVM models studied in this part. Showing what we have commented, the better performance of SVM with Radial Kernel:

```{r comparative-svm, echo=FALSE}
all_svm_results <- bind_rows(
  evaluate_model(test_data$exited, svc_probs, "SVC Linear (Cost-Sens 0.09)", 0.09),
  evaluate_model(test_data$exited, rbf_probs, "SVM RBF (Cost-Sens 0.09)",    0.09)
)

knitr::kable(all_svm_results,
             caption = "Comparative evaluation of SVM models")
```

We now place the SVM results alongside the best probabilistic models built in the previous project, all evaluated under the cost-sensitive threshold $\pi^* = 0.09$.

```{r full-comparison-prep, echo=FALSE}
# Reference results from Midterm 1 (reproduced from the report tables)
midterm1_results <- tribble(
  ~Model,                        ~Threshold, ~AUC,  ~Accuracy, ~Sensitivity, ~Specificity, ~Precision, ~F1,   ~Exp_Cost,
  "QDA (No Resampling)",         0.09,       0.817, 0.608,     0.894,        0.535,        0.330,      0.482, 5.853,
  "Naive Bayes (Oversampling)",  0.09,       0.839, 0.441,     0.961,        0.308,        0.262,      0.412, 6.313,
  "LDA (SMOTE)",                 0.09,       0.779, 0.248,     0.998,        0.056,        0.213,      0.351, 7.569,
  "Logistic Regression (SMOTE)", 0.09,       0.779, 0.240,     0.998,        0.046,        0.211,      0.348, 7.644
)

classification_results <- bind_rows(
  knn_eval %>% mutate(Method = "KNN (k=6 Oversampling)"),
  tree_results %>% mutate(Method = "Tree (Pruned Oversampling)"),
  bagging_up %>% mutate(Method = "Bagging (Oversampling)"),
  rf_up %>% mutate(Method = "Random Forest (Oversampling)"),
  gbm_cls_up %>% mutate(Method = "GBM (Oversampling)"),
  xgb_up %>% mutate(Method = "XGBoost (Oversampling)")
) %>%
  dplyr::select(-Best_Tree_Size, -Threshold_Used, -Balanced_Accuracy)
```

```{r full-comparison-table, echo=FALSE}
classification_results_clean <- classification_results %>%
  dplyr::rename(
    Model = Method,
    F1 = F1_Score,
    Exp_Cost = Expected_Cost
  ) %>%
  dplyr::mutate(
    Threshold = 0.09
  ) %>%
  dplyr::select(
    Model, Threshold, AUC, Accuracy, Sensitivity,
    Specificity, Precision, F1, Exp_Cost
  )

full_comparison <- bind_rows(
  all_svm_results,
  midterm1_results,
  classification_results_clean
) %>%
  dplyr::arrange(Exp_Cost)

knitr::kable(
  full_comparison,
  digits = 3,
  caption = "Full model comparison at threshold 0.09"
)
```

```{r comparison-plot, echo=FALSE}
full_comparison %>%
  select(Model, AUC, Sensitivity, Specificity, Exp_Cost) %>%
  pivot_longer(c(AUC, Sensitivity, Specificity, Exp_Cost),
               names_to = "Metric", values_to = "Value") %>%
  mutate(
    Model  = factor(Model, levels = rev(full_comparison$Model)),
    Metric = factor(Metric, levels = c("AUC","Sensitivity","Specificity","Exp_Cost"))
  ) %>%
  ggplot(aes(Model, Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.9) +
  coord_flip() +
  scale_fill_viridis_d(option = "E") +
  facet_wrap(~Metric, scales = "free_x", nrow = 1) +
  labs(
    title    = "Full Model Comparison at Cost-Sensitive Threshold (π* = 0.09)",
    x = "", y = "Score"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8))
```

The full comparison leads to these main conclusions.

First, the **linear SVC** performs clearly worse than the best non-linear models. Its AUC remains well below that of QDA, Naive Bayes and the RBF SVM, which confirms that a purely linear decision boundary is too restrictive for this dataset. In this sense, the linear SVM works mainly as a useful benchmark rather than as a competitive final model.

Second, the **RBF SVM** substantially improves on the linear SVM and becomes the strongest SVM specification in the project. Its AUC reaches **0.835**, which is the highest among all models except Naive Bayes, and slightly above QDA (**0.817**). This suggests that the RBF kernel captures relevant non-linear structure that the linear models miss.

Third, under the specific asymmetric cost structure adopted in the project, **QDA remains the best operational model**, with the lowest expected cost (**5.853**). The RBF SVM is still very competitive (**6.478**) and clearly outperforms the linear SVM, but it does not surpass QDA once the final classification rule is evaluated at the cost-sensitive threshold. This means that, although the RBF SVM ranks observations slightly better overall, that advantage does not fully translate into the lowest cost under the particular threshold and cost weights used here.

The **pruned tree** stands out as one of the simplest and most interpretable models, and it even achieves a competitive AUC compared with models such as KNN or logistic regression. However, this advantage does not translate into better sensitivity or lower expected cost. Therefore, the tree offers greater interpretability, but at the expense of a worse performance in our main business metric.

Overall, the comparison reinforces the main message of both parts of the project: this is a difficult classification problem, with substantial overlap between churners and non-churners and limited gains from simple linear boundaries. Non-linear methods are clearly preferable, but even among them, no model dominates on every metric. For the current cost structure, the Random Forest model with $mtry = 3$ appears to be the most suitable choice for our business case because it minimizes expected cost and achieves very high sensitivity, while the RBF SVM stands out as the best SVM alternative and a strong competitor in terms of overall discrimination.


Would you like me to review the final PyTorch Neural Network section next?